\large{\bf Large effects and the infinitesimal model }

### 2023-07-08 writeup.tex
Various edits.

\DeclareMathOperator*{\Log}{Log}
\DeclareMathOperator*{\Cauchy}{Cauchy}
\DeclareMathOperator*{\Bernoulli}{Bernoulli}
\DeclareMathOperator*{\sgn}{sgn}

@@

\author{ \begin{small}
    Todd Parsons \&
    Peter Ralph
\end{small}}

@@

Since $\varphi$ and $\Phi$ are characteristic functions, they are absolutely continuous and $\phi(0) = 1$, and thus non-zero in a neighbourhood of 0. In particular, $\psi(\theta) = \Log{\varphi}(\theta)$ and $\Psi(\theta) = \Log{\Phi(\theta)}$ exist in some neighbourhood of 0 as well\footnote{We use $\Log$ to denote the principal value of the complex logarithm, for which the imaginary part always lies in the interval $(-\pi,\pi]$.}.  From the above, we have
\[
	\Log{F^{(n)}(\varphi)(\theta)} = {\textstyle 2^{n} \psi\left(\frac{\theta}{2^{n}}\right)} 
		+ \sum_{k=1}^{n-1}  {\textstyle 2^{k} \Psi\left(\frac{\theta}{2^{k}}\right)}.
\]
We briefly explore what possible limiting values are valid characteristic functions.  We refer to the first term on the left hand side as the \emph{reproduction term} and the remaining sum as the \emph{noise term}, and only consider to the case when both have well-posed limits independent of the other.

\paragraph{Reproductive Terms:}
First consider the limit
\[
		\alpha(\theta) = \lim_{n \to \infty} 2^{n}\psi\left(\frac{\theta}{2^{n}}\right).
\]
Because $\psi(\theta)$ is the logarithm of a characteristic function, it is continuous in some neighbourhood of, and vanishes at 0.
If $\phi$ is not differentiable at zero (and so the corresponding distribution
does not have a well-defined mean),
the limit does not exist.
On the other hand if $\psi$ is differentiable at zero,
then $\alpha$ is continuous there,
and for all integers $k$, 
\[
	\alpha(2^{k} \theta) = \lim_{n \to \infty} 2^{n}\psi\left(\frac{2^{k} \theta}{2^{n}}\right) =  \lim_{n \to \infty} 2^{k} 2^{n-k} \psi\left(\frac{\theta}{2^{n-k}}\right) = 2^{k} \alpha(\theta).
\]
In particular, we see that given any continuous function $\alpha(\theta)$, $\theta \in [1,2]$, such that $\alpha(2) = 2\alpha(1)$, $\alpha$ is defined for all positive reals.  We can extend the definition to the negative reals via the Hermitian property for $\phi(\theta)$: $\phi(-\theta) = \overline{\phi(\theta)}$ implies that $\alpha(-\theta) = \overline{\alpha(\theta)}$.

Now, from the above, 
\[
	\frac{\alpha(\theta)}{\theta} = \frac{\alpha(2^{k} \theta)}{2^{k}\theta} \to \alpha'(0+),
\]
as $k \to - \infty$ if the latter exists (\ie there is a unique right-hand limit at zero; this is the case if \eg we require $\alpha(\theta)$, and thus $\phi(\theta)$, to be convex functions).  In this case, setting $-\gamma + i m = \alpha'(0+)$, we have
\[
	\alpha(\theta) = \begin{cases} 
		i m\theta - \gamma \theta & \text{if $\theta \geq 0$, and}\\
		i m\theta + \gamma \theta & \text{if $\theta < 0$}.
	\end{cases}
\]
Now, if $\phi(\theta)$ is the characteristic function of a random variable $Z$, Jensen's inequality tells us that for all $\theta$,
\[
	|\phi(\theta)| = \left|\E\left[e^{i\theta Z}\right]\right| \leq \E\left[\left|e^{i\theta Z}\right|\right] = 1
\]
so $\gamma \geq 0$. If $\gamma > 0$, then we recognize $e^{\alpha(\theta)} = e^{im \theta -\gamma|\theta|}$ as the characteristic function of a Cauchy distribution with median 0 and mean absolute deviation $\gamma$, whereas if $\gamma = 0$, then $e^{\alpha(\theta)} = e^{im \theta}$ is the Fourier transform of a Dirac mass at $m$, \ie the characteristic function for a monomorphic population with all individuals having trait value equal to $m$.

@@

\paragraph{Noise Terms}
We now turn our attention to the noise terms.  A necessary condition for $\sum_{k=1}^{\infty} 2^{k} \Psi\left(\frac{\theta}{2^{k}}\right)$ to converge is that $2^{k} \Psi\left(\frac{\theta}{2^{k}}\right) \to 0$; since $\Psi(0) = 0$, this implies that if it exists, $\Psi'(0+) = 0$.  In particular, as we would expect, if $R$ has a mean, it must be 0.  If, on the other hand, $\Psi(\theta)$ is twice differentiable, so $R$ has mean 0 and finite variance, say $\Psi''(0) = \eta^{2}$.  Taylor's theorem then tells us that for $\theta$ sufficiently small, 
\[
	\Psi(\theta) \frac{\eta^{2}}{2} \theta^{2} + r(\theta)\theta^{2},
\]
where $r(\theta) \to 0$ as $\theta \to 0$.  Thus,
\begin{equation*}
	\frac{\textstyle 2^{k+1} \Psi\left(\frac{\theta}{2^{k+1}}\right)}{\textstyle 2^{k} \Psi\left(\frac{\theta}{2^{k}}\right)}
	= \frac{1}{2} \frac{\eta^{2} \theta^{2} + {\textstyle  r\left(\frac{\theta}{2^{k+1}}\right)}\theta^{2}}{\eta^{2} \theta^{2} + {\textstyle  r\left(\frac{\theta}{2^{k}}\right)}\theta^{2}}
	\to \frac{1}{2},
\end{equation*}
as $k \to \infty$ and the sum converges by the ratio test.  In this case,
\begin{align*}
	 \sum_{k=1}^{n-1}  {\textstyle 2^{k} \Psi\left(\frac{\theta}{2^{k}}\right)}
	 &=  \sum_{k=1}^{n-1} 2^{k} {\textstyle \left(- \frac{\eta^{2}}{2}  \frac{\theta^{2}}{2^{2k}}
		+ r\left(\frac{\theta}{2^{k}}\right)\frac{\theta^{2}}{2^{2k}}\right)}\\
	&=- \eta^{2} \theta^{2} (1 - 2^{-n+1})  + \sum_{k=1}^{n-1} {\textstyle r\left(\frac{\theta}{2^{k}}\right)\frac{\theta^{2}}{2^{k}}}
\end{align*}
If we assume that $r(\theta) \equiv 0$, so that the noise is $N(0,\eta^{2})$, then the limiting distribution is $N(0,2\eta^{2})$. Note that when  $r(\theta)$ is non-zero, we don't expect the higher order terms to vanish.

It is unclear what more can be said in general. If however, we specialize to the case when $R$ is a stable law, we can explicitly compute the limiting noise term.  We recall that if $\Psi(\theta)$ corresponds to a stable law (see \eg \cite{breiman1992probability}), then
\[
	\Psi(\theta) = im\theta + c|\theta|^{\alpha} (1-i\beta \sgn(\theta)\omega_{\alpha}(\theta))
\]
for $\alpha \in (0,2]$, $\beta \in [-1,1]$, $c \geq 0$, and $m \in \mathbb{R}$ and
\[
	\omega_{\alpha}(\theta) = \begin{cases}
		\tan{\frac{\pi\alpha}{2}} & \text{if $\alpha \neq 1$, and}\\
		-\frac{2}{\pi}\log{|\theta|} & \text{if $\alpha = 1$.}
	\end{cases}
\]
By our general remarks above, we can assume $m = 0$ and exclude the case $\alpha = 1$, $\beta = 0$, for which  $\Psi'(0+) = c$. We then have
\begin{align*}
	 \sum_{k=1}^{n-1}  {\textstyle 2^{k} \Psi\left(\frac{\theta}{2^{k}}\right)}
	 &=  \sum_{k=1}^{n-1} 2^{(1-\alpha)k} c|\theta|^{\alpha} {\textstyle\left(1-i \beta\sgn(\theta)\left(\frac{\theta}{2^{k}}\right)
	 	 \omega_{\alpha}\left(\frac{\theta}{2^{k}}\right)\right)}\\
	&= \begin{cases}
		 c|\theta|^{\alpha}\frac{1-2^{(1-\alpha)n}}{1-2^{1-\alpha}}\left(1-i\beta \sgn(\theta) \tan{\frac{\pi\alpha}{2}}\right) & \text{if $\alpha \neq 1$, and}\\	
         (n-1) c|\theta|^{\alpha}  \left(1+\frac{2i\beta}{\pi}\sgn(\theta)\left(\log{|\theta|} - \frac{n}{2} \log{2}\right)\right) & \text{if $\alpha = 1$,}
	\end{cases}
\end{align*}
which converges as $n \to \infty$ if and only if $\alpha > 1$, in which case the limit is 
\[
	 \frac{c}{1-2^{1-\alpha}}|\theta|^{\alpha}\left(1-i\beta \sgn(\theta) \tan{\frac{\pi\alpha}{2}}\right), 
\]
which corresponds to a $\left(\alpha,\beta, \frac{c}{1-2^{1-\alpha}},0\right)$-stable law.

@@

\begin{figure}
    \begin{center}
        \includegraphics{sims/neutral_seg_noise_small}
    \end{center}
    \caption{
        As in Figure~\ref{fig:seg_noise},
        except computed using only mutations with absolute value effect size less than 0.2.
        The same quantities for the simulations of Figure~\ref{fig:sel_seg_noise}
        with stabilizing selection are shown in Figure~\ref{fig:sel_seg_noise_small}.
        \label{fig:seg_noise_small}
    }
\end{figure}

@@

\subsection*{Acknowledgements}
Thanks go to Nate Pope for useful discussion
and to Gregor Gorjanc for insight into modern uses
of the animal model and genomic selection.

@@

\begin{figure}
    \begin{center}
        \includegraphics{sims/selected_seg_noise_small}
    \end{center}
    \caption{
        As in Figure~\ref{fig:seg_noise},
        except computed using only mutations with absolute value effect size less than 0.2.
        \label{fig:sel_seg_noise_small}
    }
\end{figure}

>>>>

\DeclareMathOperator*{\Log}{Log}
\DeclareMathOperator*{\Cauchy}{Cauchy}
\DeclareMathOperator*{\Bernoulli}{Bernoulli}
\DeclareMathOperator*{\sgn}{sgn}
\DeclareMathOperator*{\re}{Re}

@@

\author{ \begin{small}
    Todd L. Parsons (LPSM, Sorbonne Universit\'e, CNRS UMR 8001, Paris, France, 75005) \&
    Peter Ralph
\end{small}}

@@

Since $\varphi$ and $\Phi$ are characteristic functions, they are absolutely continuous and $\phi(0) = \Phi(0) = 1$, and thus both are non-zero in a neighbourhood of 0. In particular, $\psi(\theta) = \Log{\varphi}(\theta)$ and $\Psi(\theta) = \Log{\Phi(\theta)}$ exist in some neighbourhood of 0 as well\footnote{We use $\Log$ to denote the principal value of the complex logarithm, for which the imaginary part always lies in the interval $(-\pi,\pi]$.}.  From the above, we have
\[
	\Log{F^{(n)}(\varphi)(\theta)} = {\textstyle 2^{n} \psi\left(\frac{\theta}{2^{n}}\right)} 
		+ \sum_{k=1}^{n-1}  {\textstyle 2^{k} \Psi\left(\frac{\theta}{2^{k}}\right)}.
\]
We briefly explore what possible limiting values are valid characteristic functions.  We refer to the first term on the left hand side as the \emph{reproduction term} and the remaining sum as the \emph{noise term}, and only consider to the case when both have well-posed limits independent of the other.

\paragraph{Noise Terms}
First, consider the noise terms.  A necessary condition for $\sum_{k=1}^{\infty} 2^{k} \Psi\left(\frac{\theta}{2^{k}}\right)$ to converge is that $2^{k} \Psi\left(\frac{\theta}{2^{k}}\right) \to 0$; since $\Psi(0) = 0$, this implies that if it exists, $\Psi'(0+) = 0$.  In particular, as we would expect, if $R$ has a mean, it must be 0.  If, on the other hand, $\Psi(\theta)$ is twice differentiable, so $R$ has mean 0 and finite variance, say $\Psi''(0) = \eta^{2}$.  Taylor's theorem then tells us that for $\theta$ sufficiently small, 
\[
	\Psi(\theta) \frac{\eta^{2}}{2} \theta^{2} + r(\theta)\theta^{2},
\]
where $r(\theta) \to 0$ as $\theta \to 0$.  Thus,
\begin{equation*}
	\frac{\textstyle 2^{k+1} \Psi\left(\frac{\theta}{2^{k+1}}\right)}{\textstyle 2^{k} \Psi\left(\frac{\theta}{2^{k}}\right)}
	= \frac{1}{2} \frac{\eta^{2} \theta^{2} + {\textstyle  r\left(\frac{\theta}{2^{k+1}}\right)}\theta^{2}}{\eta^{2} \theta^{2} + {\textstyle  r\left(\frac{\theta}{2^{k}}\right)}\theta^{2}}
	\to \frac{1}{2},
\end{equation*}
as $k \to \infty$ and the sum converges by the ratio test.  In this case,
\begin{align*}
	 \sum_{k=1}^{n-1}  {\textstyle 2^{k} \Psi\left(\frac{\theta}{2^{k}}\right)}
	 &=  \sum_{k=1}^{n-1} 2^{k} {\textstyle \left(- \frac{\eta^{2}}{2}  \frac{\theta^{2}}{2^{2k}}
		+ r\left(\frac{\theta}{2^{k}}\right)\frac{\theta^{2}}{2^{2k}}\right)}\\
	&=- \eta^{2} \theta^{2} (1 - 2^{-n+1})  + \sum_{k=1}^{n-1} {\textstyle r\left(\frac{\theta}{2^{k}}\right)\frac{\theta^{2}}{2^{k}}}
\end{align*}
If we assume that $r(\theta) \equiv 0$, so that the noise is $N(0,\eta^{2})$, then the limiting distribution is $N(0,2\eta^{2})$. Note that when  $r(\theta)$ is non-zero, we don't expect the higher order terms to vanish.

It is unclear what more can be said in general. If however, we specialize to the case when $R$ is a stable law, we can explicitly compute the limiting noise term.  We recall that if $\Psi(\theta)$ corresponds to a stable law (see \eg \cite{breiman1992probability}), then
\[
	\Psi(\theta) = im\theta + c|\theta|^{\alpha} (1-i\beta \sgn(\theta)\omega_{\alpha}(\theta))
\]
for $\alpha \in (0,2]$, $\beta \in [-1,1]$, $c \geq 0$, and $m \in \mathbb{R}$ and
\[
	\omega_{\alpha}(\theta) = \begin{cases}
		\tan{\frac{\pi\alpha}{2}} & \text{if $\alpha \neq 1$, and}\\
		-\frac{2}{\pi}\log{|\theta|} & \text{if $\alpha = 1$.}
	\end{cases}
\]
By our general remarks above, we can assume $m = 0$ and exclude the case $\alpha = 1$, $\beta = 0$, for which  $\Psi'(0+) = c$. We then have
\begin{align*}
	 \sum_{k=1}^{n-1}  {\textstyle 2^{k} \Psi\left(\frac{\theta}{2^{k}}\right)}
	 &=  \sum_{k=1}^{n-1} 2^{(1-\alpha)k} c|\theta|^{\alpha} {\textstyle\left(1-i \beta\sgn(\theta)\left(\frac{\theta}{2^{k}}\right)
	 	 \omega_{\alpha}\left(\frac{\theta}{2^{k}}\right)\right)}\\
	&= \begin{cases}
		 c|\theta|^{\alpha}\frac{1-2^{(1-\alpha)n}}{1-2^{1-\alpha}}\left(1-i\beta \sgn(\theta) \tan{\frac{\pi\alpha}{2}}\right) & \text{if $\alpha \neq 1$, and}\\	
         (n-1) c|\theta|^{\alpha}  \left(1+\frac{2i\beta}{\pi}\sgn(\theta)\left(\log{|\theta|} - \frac{n}{2} \log{2}\right)\right) & \text{if $\alpha = 1$,}
	\end{cases}
\end{align*}
which converges as $n \to \infty$ if and only if $\alpha > 1$, in which case the limit is 
\begin{equation}\label{eq:stablenoises}
	 \frac{c}{1-2^{1-\alpha}}|\theta|^{\alpha}\left(1-i\beta \sgn(\theta) \tan{\frac{\pi\alpha}{2}}\right), 
\end{equation}
which corresponds to a $\left(\alpha,\beta, \frac{c}{1-2^{1-\alpha}},0\right)$-stable law.

\paragraph{Reproductive Terms:}
We now turn our attention to the limit
\[
		\alpha(\theta) = \lim_{n \to \infty} 2^{n}\psi\left(\frac{\theta}{2^{n}}\right)
\]
If it exists, then for all integers $k$, 
\[
	\alpha(2^{k} \theta) = \lim_{n \to \infty} 2^{n}\psi\left(\frac{2^{k} \theta}{2^{n}}\right) =  \lim_{n \to \infty} 2^{k} 2^{n-k} \psi\left(\frac{\theta}{2^{n-k}}\right) = 2^{k} \alpha(\theta).
\]
and, since $\alpha(0) = 0$,
\[
	\frac{\alpha(\theta)}{\theta} = \frac{\alpha(2^{k} \theta)}{2^{k}\theta} 
	\to \begin{cases} 
		\alpha'(0+) & \text{if $\theta > 0$, and}\\
		\alpha'(0-) & \text{if $\theta < 0$,}
	\end{cases}
\]
as $k \to - \infty$ if the limits exist (\ie there are unique left-and right-hand limit at zero; this is the case if \eg we require $\alpha(\theta)$, and thus $\phi(\theta)$, to be convex functions).  We henceforth assume this.

Since $\varphi(\theta)$ is a characteristic function, it has the Hermitian property: $\phi(-\theta) = \overline{\phi(\theta)}$, and thus $\psi(-\theta) = \overline{\psi(\theta)}$ and $\alpha(-\theta) = \overline{\alpha(\theta)}$.  In particular, if one (and by extension both) limits exist, we must have $\alpha'(0-) = \overline{\alpha'(0+)}$.  In this case, setting $-\gamma + i m = \alpha'(0+)$, we have
\[
	\alpha(\theta) = \begin{cases} 
		i m\theta - \gamma \theta & \text{if $\theta \geq 0$, and}\\
		i m\theta + \gamma \theta & \text{if $\theta < 0$}.
	\end{cases}
\]
If $\gamma > 0$, then we recognize $e^{\alpha(\theta)} = e^{im \theta -\gamma|\theta|}$ as the characteristic function of a Cauchy distribution with median 0 and mean absolute deviation $\gamma$, whereas if $\gamma = 0$, then $e^{\alpha(\theta)} = e^{im \theta}$ is the Fourier transform of a Dirac mass at $m$, \ie the characteristic function for a monomorphic population with all individuals having trait value equal to $m$.    In general, it does not seem possible to exclude $\gamma < 0$. However, recalling that if $\phi(\theta)$ is the characteristic function of a random variable $Z$, Jensen's inequality tells us that for all $\theta$,
\begin{equation}\label{eq:CF-J}
	|\phi(\theta)| = \left|\E\left[e^{i\theta Z}\right]\right| \leq \E\left[\left|e^{i\theta Z}\right|\right] = 1,
\end{equation}
we can conclude $\gamma \geq 0$ when the limiting noise term is a mixture of Gaussians and stable laws of the form \eqref{eq:stablenoises}: we then have 
\[
	|\phi(\theta)| = e^{\re(\alpha(\theta) + o(|\theta|))} = e^{\gamma |\theta| + o(|\theta|))},
\] 
and thus is less than or equal to one in a small neighbourhood of $0$ if and only if $\gamma < 0$.

@@

@@

\begin{figure}
    \begin{center}
        \includegraphics{sims/neutral_seg_noise_small}
    \end{center}
    \caption{
        As in Figure~\ref{fig:seg_noise},
        except computed using only mutations with absolute value effect size less than 0.1.
        The same quantities for the simulations of Figure~\ref{fig:sel_seg_noise}
        with stabilizing selection are shown in Figure~\ref{fig:sel_seg_noise_small}.
        \label{fig:seg_noise_small}
    }
\end{figure}

@@

\subsection*{Acknowledgements}
Thanks go to Nate Pope for useful discussion
and to Gregor Gorjanc for insight into modern uses
of the animal model and genomic selection.  This project was started on a visit by TLP to the University of Oregon that was partially supported by the CNRS PEPS grant ``Jeunes Chercheuses et Jeunes Chercheurs''.

@@

\begin{figure}
    \begin{center}
        \includegraphics{sims/selected_seg_noise_small}
    \end{center}
    \caption{
        As in Figure~\ref{fig:seg_noise},
        except computed using only mutations with absolute value effect size less than 0.1.
        \label{fig:sel_seg_noise_small}
    }
\end{figure}
